---
title: 'Intermediate R: Statistics for Graduate Students'
author: "Nnenna Asidianya"
date: "2/26/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

# R Markdown

Remove the hashtag # to install R markdown to follow along

```{r}
#install.packages("rmarkdown")
```

# Conditionals and Control Flow

## Equality (or not)

```{r}
#Example 1: Equality 

x=3
y=4
x==y

#Example 2: Not equality 

x!=y

#Example 3: Equality of strings. 

user1<-c("userR", "user", "UserR")
user2<-c("userR", "useR", "UserR")
user1==user2 #this is a logical vector

#Bonus operation; logical operations for data manipulation

user<-ifelse(user1==user2, "userR", "userR")
```

## Greater than or less than 

```{r}
library(tidyverse)
height <- c(150, 145, 150, 155, 170, 180, 177)#in cm
weight <- c(60, 55, 90, 54, 75, 80, 76 ) #in kg

#this is a hacky way of doing this. There are better methods below.
features<-tibble(height, weight)
ggplot(features, aes(weight, height))+geom_point()+geom_smooth(method = "lm", se=F)

attach(features)

features2<- features[ which(weight<90), ]
```
Note: This is an inefficient way to subset the outliers in data. In general I would most likely use 'filter' tidyverse (more on this in another workshop). However, I will come back to the subset function later on. 

  
## The  & and  | operators (and and or)

```{r}
#recall
features

#weight that is equal to 90 and height that is equal to 150
outlier <-features[ which(weight==90 & height==150), ]
outlier

#weight that is equal to 90 and height that is equal to 150
features3<-features[ which(weight==90 | height==150), ]
features3

```
# Loops

## For loop

```{r}
set.seed(1)

rv<-rnorm(1000)

usq<-matrix("NA", 100, 1)


for(i in 1:100) {
  
  usq[i] <- rv[i]*rv[i]
  #print(usq[i])
}

usq<-data.frame(usq=as.numeric(unlist(usq)))
attach(usq)
ggplot(usq, aes(x=usq))+geom_histogram(col="black", fill="red")+xlab("x")+ylab("frequency")

```

## While loop 



```{r}
set.seed(1)


usq<-matrix("NA", 100, 1)

n = length(usq)
x=0

while (x<=n) {
  x<-sum(x, 1)
  i<-sum(i, 1)
  usq[i] <- rv[i]*rv[i]
  #print(usq[i])
}

usq<-data.frame(usq=as.numeric(unlist(usq)))
attach(usq)
ggplot(usq, aes(x=usq))+geom_histogram(col="black", fill="red")+xlab("x")+ylab("frequency")

```



```{r}
weight
height

# Code the for loop with conditionals
for (i in weight) {
  if (i > 80) {
    print("You're an outlier!")
  }
  else {
    print("Nothing to see here!")
  }
  print(i)
}
```

# What is a function in R?


An R function is created by using the keyword *function*. The basic syntax of an R function definition is as follows:


```{r}
function_name <- function(arg_1,..) {
   Function_body 
}
```

## Example 1: Perfect squares 

```{r}
#Create a function to print out the values for perfect squares. 
perfect.squares<-function(x){for (x in 1:10){
  if(x%%x == 0)
    b<-x^2
  print (b)
}}

#pass a value to the function
perfect.squares(10)
```

## Example 2: Processing course grades

```{r}
#Create a function with more than one argument
translate<-function(x,y){
  if (x=="NA"){x=0}
  else if (x!="NA"){x=x}
  print(x)
  if (y== -1){y="NA"}
  else if (y!=-1){y=y}
  print(y)

}


#Testing the function with input values 
x<-"NA"
y<-"-1"

x2<-1
y2<-0

translate(x, y)
translate(x2, y2)
```


## Ifelse function in R

In order to understand the brilliance of this function let's go back to the toy example 2.

```{r}
x3<-c(0, "NA", 1)
y3<-c(-1, "NA", 80)

translate(x3,y3)
```


The `ifelse` is a built in base R function that returns a value with the same length as the test, rather than the evaluation at the first element. The elements selected from either `yes` or `no` depend on whether the `yes` condition is true or false. 

```{r}
z<-x3
ifelse(z==0, "NA", z)
```
More on this a bit later when I get to subsetting and/or data manipulation in R. 


#Subset function and tidy data

I am developing an R package for grade processing:

```{r}
#Base R
#data = read.csv("StatisticsGrades.csv")

#Tlibrary(tidyverse)
#data = read_csv("StatisticsGrades.csv")
#attach(data)

#This will not be runnnable on your end so use the above vode. 
library(myPackage)
attach(Grades)

head(Grades)

TT1 = `Term Test 1 (/505)` 
TT2= `Term Test 2 (/57)`


```

The data set I will work with is inside of the package I am developing but is currently not available on Git. 

```{r}

dim(Grades)

#change the name to data for your own data frame from Grades
newdata <- subset(Grades, TT1 !=-1 | TT2!= -1,)
dim(newdata)


newgrades <- subset(Grades, TT1 !=-1 & TT2!= -1,)
dim(newgrades)

```


The first subset data scanned students that had either a TT1 or TT2 with a "-1" indicator. The second subset data set looked for students who had both TT1 and TT2 grades missing. 

## new.function in myPackage

```{r, eval=F}
TT1_new = new.grades(TT1)
TT1_new

TT2_new = new.grades(TT2)

TT_new = subset(Grades, TT1_new!="NA" & TT2_new!="NA")
dim(TT_new)

TT_nu<-as.numeric(unlist(TT_new))
```

I have gone further than removing "-1". Any "NA"s in the data set. I converted those students without a noted absence to a value of "0." It turns out that there are no true 0's in this data set. 

## Tidyverse 

Is there an asssociation between TT1 grades and TT2 grades? You can refer to the data set called 'newgrades.'

```{r}
#install.packages("tidyverse")
library(tidyverse)

newgrades<-newgrades %>% mutate(TT1=`Term Test 1 (/505)`, TT2=`Term Test 2 (/57)`)
attach(newgrades)

ggplot(newgrades, aes(x=TT1, y=TT2))+geom_point()+geom_smooth(method="lm")
```

